---
title: "Stroke_Prediction_Analysis_Group_C"
format: html
editor: visual
---

## 1. Introduction and Data Structure

### 1.1 Overview & Motivation

### 1.2 Related Work

### 1.3 Research Questions

### 1.4 Data Source

### 1.5 Data Import

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
library(tidyverse)
```

You can add options to executable code like this

```{r}
#import of the dataset
stroke_db <- read.csv('C:/Users/fabio/OneDrive/Semester 1/Data Science in Business Analytics/Project/stroke_dataset.csv')

#saving as a tibble
stroke_tb <- as_tibble(stroke_db)
stroke_tb
```

In total, there are 12 variables including our binary target variable 'Stroke', which indicates whether the person either had a stroke (1) or not (1).

```{r}

```

## 2. Data Cleaning & Wrangeling

### 2.1 Missing Values

The features in our dataset are fairly clean. The only column with missing values is BMI.

### 2.3 Outlier Analysis

### 2.4 Feature Scaling

## 3. Exploratory Data Analysis

## 4. Predictive Model

## 5. Model Evaluation and Predictions

## 6. Conclusions
