---
title: "Understanding Stroke: A Data-Driven Approach"
format: revealjs
editor: visual
---

```{r, warning=FALSE}
library(naniar)
library(tidyverse)
library(caret)
library(plotly)
library(knitr)
```

## Stroke: A Global Health Emergency

The Problem:

```         
•   15M strokes/year (WHO): 5M deaths, 5M disabilities.
•   Stroke disrupts brain blood flow: caused by blockage or bleeding.
```

Project Goals:

```         
•   Predict stroke risk using data.
•   Analyze key factors like age, heart disease, & lifestyle.
•   Visualize insights for better understanding.
```

Key Question:

What factors best predict stroke, and how can we intervene?

## Dataset Overview

Data Source

```         
    •   Origin: Found on Kaggle.com. Owned and last updated in 2020 by Federico Soriano Palacios.
    •   Observations: 5110
    •   Features: 11
```

Key Variables

```         
    •   Target Variable: stroke (0 = No, 1 = Yes)
    •   Demographics: gender, age, Residence_type
    •   Health Metrics: hypertension, heart_disease, avg_glucose_level, bmi
    •   Lifestyle: smoking_status, work_type, ever_married
```

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Original Dataset
#kable(stroke_tb)

```

## Data Transformations

Before preprocessing:

```         
Dataset is fairly clean
    •   No duplicate rows.
    •   Missing values in bmi and smoking_status.
```

After data transformations:

```         
Variable Adjustments:
•   Reclassified variables: bmi → numerical; hypertension, heart_disease → categorical.
•   Removed irrelevant column: id.

Handling Missing Values:
•   Replaced missing values: bmi → median, smoking_status → mode.

Outlier Analysis:
•   Visualized outliers in age, avg_glucose_level, and bmi using boxplots.
•   Outliers Identified:
•   avg_glucose_level: 12.25% of dataset (627 points).
•   bmi: 2.47% of dataset(126 points).
•   Retained outliers for now due to potential relevance as risk factors.

Categorical Variables:
•   Removed infrequent categories in gender (Other) and work_type (Never_worked).
•   Noted rare occurrences in heart_disease, hypertension, and stroke.
```

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
#colSums(stroke_tb == "Unknown" | stroke_tb == "N/A" | stroke_tb == "")
```

## Findings and Results

### Predictive model

## Model evaluation

## Conclusion

## Requirements:

Introduction: Briefly introduce your project goals and motivation.

Research Questions: What were the key research questions?

Data: Provide a high-level overview of your data sources and preprocessing.

Findings and Results: Highlight your key findings, insights, or model results.

Conclusion: Summarize your main takeaways and future potential directions.

Length: 7 minutes or less. The time limit will be strictly enforced. Content:

Tell your story well! Focus on delivering clear, concise, and engaging insights from your project. 🚀
