---
format: html
---
# Data

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(naniar)
library(tidyverse)
library(caret)
library(plotly)
library(knitr)
library(kableExtra)
library(dplyr)
```

## Data Source

The data utilized for this project originates from the [**Stroke Prediction Dataset**](https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset/data) available on [**Kaggle**](https://www.kaggle.com/), which is a public platform that hosts a variety of datasets curated by data scientists and researchers. This specific dataset includes health-related variables that are relevant for predicting occurrences of strokes based on features like age, gender, and medical history. Unfortunately, we do not have more information on how the data was collected. The source is confidential and the use of it is only permitted for educational purposes.

The owner of the dataset is Federico Soriano Palacios (found thanks to the link to his LinkedIn on his Kaggle profile). The last update of the dataset was 4 years ago.

## Data Structure

We start by loading the dataset and displaying the first few rows to get an overview of the data structure. The dataset contains 5110 observations and 12 variables, including the target variable `stroke`.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Import the dataset
stroke_tb <- read.csv("../data/datasets/stroke_dataset.csv") 
stroke_tb_tibble <- as_tibble(stroke_tb)

# Display the dataset
kable(stroke_tb, format = "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  scroll_box(height = "300px")
```

Our dataset contains 5110 observations with 12 attributes. Here's a summary of the main features of the dataset:

1.  `id`:

-   Type: Integer (Continuous)
-   Description: A unique identifier for each individual in the dataset

2.  `gender`:

-   Type: Character
-   Values: "Male", "Female", or "Other"
-   Description: Gender of the individual

3.  `age`:

-   Type: Double
-   Description: Age of the individual

4.  `hypertension`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: Indicates whether the individual has hypertension (high blood pressure)

5.  `heart_disease`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: Indicates whether the individual has a heart disease

6.  `ever_married`:

-   Type: Character
-   Values: "No" or "Yes"
-   Description: Indicates whether the individual has ever been married

7.  `work_type`:

-   Type: Character
-   Values: "children", "Govt_job", "Never_worked", "Private", "Self-employed"
-   Description: Type of work the individual is engaged in, where "children" signifies that the individual is a child

8.  `Residence_type`:

-   Type: Character
-   Values: "Rural" or "Urban"
-   Description: Indicates the living area of the individual

9.  `avg_glucose_level`:

-   Type: Double
-   Description: Average glucose level in the blood

10. `bmi`:

-   Type: Character (Originally), Double (After Preprocessing)
-   Description: Body Mass Index, a measure of body fat based on height and weight

11. `smoking_status`:

-   Type: Character
-   Values: "formerly smoked", "never smoked", "smokes", "Unknown"
-   Description: Indicates the smoking habits of the individual, where "Unknown" signifies that the smoking status information is not available

12. `stroke`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: The target variable indicating whether the individual has had a stroke

## Data Preprocessing

In the next steps, we examine our dataset on potential issues such as missing values, duplicated rows, and outliers. We also address the data types of the variables and prepare the dataset for further analysis.

We start by verifying that no observations are duplicated in our dataset:
```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
sum(duplicated(stroke_tb))
```

No duplicated rows are present. As we have seen before, our dataset includes both categorical and numerical variables. To prepare the dataset for the EDA and the modeling part, we clearly distinguish these types from the start.

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Identify categorical and numerical variables 
categorical_variables <- names(stroke_tb)[sapply(stroke_tb, function(x) is.character(x) || is.integer(x))]
numerical_variables <- names(stroke_tb)[sapply(stroke_tb, is.double)]

# Print the lists of categorical and numerical variables for a quick check
cat(sprintf("Categorical features: %s\nNumerical features: %s",
            paste(categorical_variables, collapse = ", "),
            paste(numerical_variables, collapse = ", ")))
```

Here, we observe that some variables don't get recognized correctly. `bmi` is a numerical feature and `hypertension` and `heart_disease` ought to be categorical. We will address these issues in the next steps.

### Missing Values

We start by identifying the number of missing values in each column:
```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Sum of missing values in each column
colSums(stroke_tb == "Unknown" | stroke_tb == "N/A" | stroke_tb == "")
```

We address missing values in the features `bmi` and `smoking_status` by replacing them with `N/A`. Additionally, we ensure that `bmi` is recognized as a numerical variable and update the lists of categorical and numerical columns, as `bmi` is now a double. The `id` column is also removed, as it is not relevant for our analysis.

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Clean the dataset
cleaned_stroke_tb <- replace_with_na(
  data = stroke_tb,
  replace = list(
    bmi = c("N/A", ""),  
    smoking_status = c("Unknown", "") 
  )
) %>%
  mutate(
    bmi = as.numeric(bmi)  # Convert bmi to numeric after handling "N/A"
  ) %>%
  select(-id)  # Remove the 'id' column

# Display the dataset with up to 10 observations per page
kable(cleaned_stroke_tb, format = "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  scroll_box(height = "300px")

# Identify categorical variables (character or factor)
categorical_variables <- names(cleaned_stroke_tb)[sapply(cleaned_stroke_tb, function(x) is.character(x) || is.integer(x))]
numerical_variables <- names(cleaned_stroke_tb)[sapply(cleaned_stroke_tb, is.double)]

# Store the vectors
categorical_vars <- categorical_variables
numerical_vars <- numerical_variables

# Print the vectors to check
cat("Categorical Variables:\n", paste(categorical_vars, collapse = ", "), "\n")
cat("Numerical Variables:\n", paste(numerical_vars, collapse = ", "), "\n")

# Save 
write.csv(cleaned_stroke_tb, "../data/datasets/cleaned_stroke_tb.csv", row.names = FALSE)
```

Then, to maintain the number of observations in our dataset, we replace the `N/A` values in `bmi` with the median and in `smoking_status` with the modal value.

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Calculate the median for bmi and the mode for smoking_status
bmi_median <- median(cleaned_stroke_tb$bmi, na.rm = TRUE)
smoking_status_mode <- names(which.max(table(cleaned_stroke_tb$smoking_status, useNA = "no")))

# Create preprocessed data by replacing NA values
preprocessed_stroke_tb <- cleaned_stroke_tb %>%
  mutate(
    bmi = ifelse(is.na(bmi), bmi_median, bmi),
    smoking_status = ifelse(is.na(smoking_status), smoking_status_mode, smoking_status)
  )

# Display the dataset with up to 10 observations per page
kable(preprocessed_stroke_tb, format = "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  scroll_box(height = "300px")

write.csv(preprocessed_stroke_tb, "../data/datasets/preprocessed_stroke_tb.csv", row.names = FALSE)
```

### Outlier Analysis

We notice that there is only three non-binary quantitative variables: `age`, `avg_glucose_level` and `bmi`. To visualize outliers in each of these variables, we use boxplots. The points outside the range defined by the whiskers were identified as outliers.

```{r}
# Define y-axis labels for each variable
y_axis_labels <- c("age" = "in years", 
                   "avg_glucose_level" = "in mg/dL", 
                   "bmi" = "in kg/m²")

# Reshape the data to long format
long_data <- preprocessed_stroke_tb %>%
  select(all_of(numerical_variables)) %>%
  pivot_longer(
    cols = everything(), 
    names_to = "Variable", 
    values_to = "Value"
  )

# Map colors to variables
variable_colors <- c("age" = "blue", "avg_glucose_level" = "orange", "bmi" = "green")

# Initialize an empty list to store individual plots
plots <- list()

# Loop through each variable and create separate boxplots
for (var in numerical_variables) {
  plot <- plot_ly(
    data = long_data %>% filter(Variable == var),
    x = ~Variable,
    y = ~Value,
    type = "box",
    name = var,
    marker = list(color = variable_colors[var])
  ) %>%
    layout(
      yaxis = list(
        title = y_axis_labels[var],  # Assign custom y-axis title
        titlefont = list(size = 14) # Optional: Adjust font size
      ),
      xaxis = list(title = "")
    )
  
  # Add the plot to the list
  plots <- c(plots, list(plot))
}

# Combine all plots into a single subplot with separate y-axes
combined_plot <- subplot(
  plots,
  nrows = 1,         # All in one row
  shareX = TRUE,     # Share x-axis
  titleX = TRUE,     # Display x-axis title
  titleY = TRUE     # Allow individual y-axes
) %>%
  layout(
    title = "Boxplots of Numerical Variables",
    showlegend = FALSE
  )

# Display the combined plot
combined_plot
```


Additionally, we sum up the outliers for each numerical variable:
```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
identify_outliers <- function(column) {
  Q1 <- quantile(column, 0.25)
  Q3 <- quantile(column, 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  # Identify Outliers
  outliers <- column[column < lower_bound | column > upper_bound]
  return(outliers)
}

# Apply the function to the variables
outliers_age <- identify_outliers(preprocessed_stroke_tb$age)
outliers_glucose <- identify_outliers(preprocessed_stroke_tb$avg_glucose_level)
outliers_bmi <- identify_outliers(preprocessed_stroke_tb$bmi)

#We create a function count_outliers that we will use to identify and count outliers in each non-binary quantitative variable 
  count_outliers <- function(column) {
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
# Count outliers
outlier_count <- sum(column < lower_bound | column > upper_bound, na.rm = TRUE)
  return(outlier_count)
}

# Create a vector to hold the outlier counts
outlier_counts <- sapply(preprocessed_stroke_tb[, c("age", "avg_glucose_level", "bmi")], count_outliers)
outlier_counts
```

Finding zero outliers for `age` indicates a nearly normal distribution among the individuals in our population. With a dataset containing 5,110 rows, the identification of 627 outliers for `avg_glucose_level`, representing approximately 12.25% of the total population, and 126 outliers for `bmi`, constituting about 2.47%. Both boxplots display a considerable number of outliers in the higher range, indicating that there are subsets of individuals in the dataset with unusually high glucose levels and BMI values. These outliers could be due to a variety of factors, such as lifestyle, health conditions, or measurement errors. Considering their importance as risk factors, we decided to keep them in the dataset for now. However, we may revisit this decision later in the analysis.

After conducting the analysis of the numerical variables, we examine the distribution of the categorical variables.
```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Count the frequency of each category in the categorical variables
lapply(preprocessed_stroke_tb[categorical_vars], table) %>%
  lapply(as.data.frame) %>%
  bind_rows(.id = "Variable") %>%
  rename(Value = Var1, Count = Freq) %>%
  kable(col.names = c("Variable", "Category", "Count"))
```

To simplify the further data exploration, we remove the less frequent categories in `gender`(Other) and `work_type`(Never_worked) as they are not representative of the majority of the population. We further notice that `heart_disease` and `hypertension` aren't very common in our dataset and `stroke` occurrences are also pretty rare, which we will need to consider in our modeling part. But more on that later in the EDA.

```{r, warning=FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Removal of observations of `Never_worked` and `Other`
outlier_stroke_tb <- preprocessed_stroke_tb %>%
  filter(work_type != "Never_worked") %>%
  filter(gender != "Other")

# save
write.csv(outlier_stroke_tb, "../data/datasets/outlier_stroke_tb.csv", row.names = FALSE)
```
