---
title: "_eda"
format: html
---

# Exploratory Data Analysis

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
library(tidyverse)
library(plotly)
library(knitr)
library(caret)
library(vcd)
library(ggplot2)
library(reshape2)
library(ggplot2)
library(plotly)
library(dplyr)
library(knitr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(scales)  # for percentage labels
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
outlier_stroke_tb <- 
  read.csv("../../data/datasets/outlier_stroke_tb.csv")
balanced_stroke_tb <-
  read.csv("../../data/datasets/balanced_stroke_tb.csv")
preprocessed_stroke_tb <-
  read.csv("../../data/datasets/preprocessed_stroke_tb.csv")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Numerical variables
numerical_variables <- c("age", "avg_glucose_level", "bmi")

# Create a tibble with the numerical variables
numerical_data <- outlier_stroke_tb %>%
  select(all_of(numerical_variables))
```

## Descriptive Statistics

We start our EDA with a summary of the numerical and categorical variables in our dataset. The summary includes the count, mean, standard deviation, minimum, 1st quartile, median, 3rd quartile, and maximum values for numerical variables.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# Generate summary of selected numerical variables
summary_table <- summary(outlier_stroke_tb[, c("age", "avg_glucose_level", "bmi")])

# Convert the summary to a data frame & display it with kable
summary_table <- as.data.frame.matrix(summary_table)
knitr::kable(summary_table, caption = "Summary of Numerical Variables", row.names = FALSE)
```

For the categorical variables, we present the total number and proportion of each category’s occurrences. We combined all categorical variables into one table for a better overview.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"
# List of categorical variables
variables_to_summarize <- c("stroke", "gender", "ever_married", "work_type", "Residence_type", "smoking_status", "hypertension", "heart_disease")

# Convert binary integer variables to factors
outlier_stroke_tb <- outlier_stroke_tb %>%
  mutate(
    stroke = as.factor(stroke),
    hypertension = as.factor(hypertension),
    heart_disease = as.factor(heart_disease)
  )

# Calculate total count for proportion calculation
total_count <- nrow(outlier_stroke_tb)

# Create summary table
summary_table <- bind_rows(
  lapply(variables_to_summarize, function(var) {
    outlier_stroke_tb %>%
      group_by(across(all_of(var))) %>%
      summarize(
        Count = n(),
        Proportion = n() / total_count,
        .groups = "drop"
      ) %>%
      rename(Type = !!sym(var)) %>%
      mutate(
        Variable = var,
        Type = as.character(Type),
        Proportion = round(Proportion * 100, 2)
      )
  })
) %>%
  select(Variable, Type, Count, Proportion) %>%
  arrange(Variable, Type)                  

# Display the combined table with kable
knitr::kable(summary_table, caption = "Summary of Categorical Variables with Proportion", row.names = FALSE)
```

## Univariate Analysis

### Numerical variables

First, we visualized all numerical variables in a histogram grouped by stroke occurrence to get an overview of the distribution of the data and their relationship with the target variable.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Function to calculate proportion of people with stroke by bins and create a line plot
create_proportion_line <- function(data, variable, stroke_variable, binwidth = 5) {
  original_data <- data[[variable]]
  stroke_data <- data[[stroke_variable]]

  # Define bin edges
  range_vals <- range(original_data, na.rm = TRUE)
  breaks <- seq(floor(range_vals[1]), ceiling(range_vals[2]) + binwidth, by = binwidth)

  # Cut the data into bins
  binned_data <- data.frame(value = original_data, stroke = stroke_data) %>%
    mutate(bin = cut(value, breaks = breaks, include.lowest = TRUE, right = FALSE))

  # Compute proportion stroke in each bin
  prop_data <- binned_data %>%
    group_by(bin) %>%
    summarize(
      n = n(),
      stroke_n = sum(stroke == 1, na.rm = TRUE),
      prop_stroke = stroke_n / n,
      .groups = "drop"
    )

  # Calculate bin midpoints
  bin_mids <- sapply(strsplit(gsub("\\[|\\]|\\(|\\)", "", levels(prop_data$bin)), ","), 
                     function(x) mean(as.numeric(x)))
  prop_data <- prop_data %>%
    mutate(bin_mid = bin_mids[as.numeric(bin)])

  # Create a line plot using plotly with the same shade of blue as the histogram
  line_plot <- plot_ly(
    data = prop_data,
    x = ~bin_mid,
    y = ~prop_stroke,
    type = 'scatter',
    mode = 'lines+markers',
    line = list(color = '#1f77b4'),
    marker = list(color = '#1f77b4')
  ) %>%
    layout(
      yaxis = list(title = "Proportion Stroke", rangemode='tozero'),
      showlegend = FALSE,
      title = paste("Proportion Stroke by", gsub("_", " ", variable))
    )

  return(line_plot)
}

# Function to create the histogram as a Plotly object
create_histogram <- function(data, variable, stroke_variable, binwidth = 5, show_legend = FALSE) {
  original_data <- data[[variable]]
  stroke_data <- data[[stroke_variable]]
  stroke_factor <- factor(stroke_data, levels = c(0, 1), labels = c("No Stroke", "Stroke"))

  plot_data <- data.frame(data = original_data, stroke = stroke_factor)

  p <- ggplot(plot_data, aes(x = data, fill = stroke)) +
    geom_histogram(binwidth = binwidth, color = "white", position = "stack") +
    labs(
      y = "Frequency",
      fill = "Stroke Status",
      title = paste("Distribution of", gsub("_", " ", variable))
    ) +
    scale_fill_manual(values = c("No Stroke" = "grey50", "Stroke" = "#1f77b4")) +
    theme_minimal() +
    theme(
      axis.title.x = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.position = ifelse(show_legend, "right", "none")
    )

  ggplotly(p) %>% layout(showlegend = show_legend)
}

# Assuming outlier_stroke_tb is defined and stroke is 0/1
line_age <- create_proportion_line(outlier_stroke_tb, "age", "stroke", binwidth = 5) %>%
  layout(xaxis = list(title = ""), yaxis = list(title = "Proportion Stroke"))
hist_age <- create_histogram(outlier_stroke_tb, "age", "stroke", binwidth = 5, show_legend = TRUE) %>%
  layout(xaxis = list(title = "Age"), yaxis = list(title = "Frequency"))

line_glucose <- create_proportion_line(outlier_stroke_tb, "avg_glucose_level", "stroke", binwidth = 5) %>%
  layout(xaxis = list(title = ""), yaxis = list(title = "Proportion Stroke"))
hist_glucose <- create_histogram(outlier_stroke_tb, "avg_glucose_level", "stroke", binwidth = 5, show_legend = FALSE) %>%
  layout(xaxis = list(title = "Average Glucose Level"), yaxis = list(title = "Frequency"))

line_bmi <- create_proportion_line(outlier_stroke_tb, "bmi", "stroke", binwidth = 5) %>%
  layout(xaxis = list(title = ""), yaxis = list(title = "Proportion Stroke"))
hist_bmi <- create_histogram(outlier_stroke_tb, "bmi", "stroke", binwidth = 5, show_legend = FALSE) %>%
  layout(xaxis = list(title = "BMI"), yaxis = list(title = "Frequency"))

# Stack line above histogram for each variable with height adjustments
age_subplot <- subplot(line_age, hist_age, nrows = 2, shareX = TRUE, titleY = TRUE, heights = c(0.3, 0.7))
glucose_subplot <- subplot(line_glucose, hist_glucose, nrows = 2, shareX = TRUE, titleY = TRUE, heights = c(0.3, 0.7))
bmi_subplot <- subplot(line_bmi, hist_bmi, nrows = 2, shareX = TRUE, titleY = TRUE, heights = c(0.3, 0.7))

# Combine all three sets into a single row
combined_subplots <- subplot(age_subplot, glucose_subplot, bmi_subplot, nrows = 1, margin = 0.05, titleX = TRUE, titleY = TRUE) %>%
  layout(title = "Histograms and Stroke Proportion by Numerical Variables")

# Display the combined subplots
combined_subplots
```


Based on the summary table for numerical variables and the histogram, we can observe the following descriptive characteristics:

-   `age`: The dataset includes individuals with ages ranging from 0.08 to 82 years, with an average age of 43.23 years but a peak around 57 years, a second peak around 78 years, and a standard deviation of 22.61 years.
-   `avg_glucose_level`: We can observe a right-skewed bimodal distribution with a mean of 106.15 and a standard deviation of 45.28. The average glucose levels range from 55.12 to 271.74 mg/dL.
-   `bmi`: The histogram shows a right-skewed distribution with a mean of 28.86 and a standard deviation of 7.85. The BMI values range from 10.3 to 97.6. This is slightly higher than the average population according to a study by the World Health Organization (WHO).

For completeness we added some violin plots of the numerical variables.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Function to create a violin plot
create_violin_plot <- function(data, variable, stroke_variable) {
  # Ensure the stroke variable is a factor
  data[[stroke_variable]] <- as.factor(data[[stroke_variable]])
  
  # Generate the violin plot with a name for the trace
  plot <- plot_ly(data, x = ~get(stroke_variable), y = ~get(variable), type = "violin",
                  box = list(visible = TRUE),
                  meanline = list(visible = TRUE),
                  name = variable) %>%  # Add this line to set the name
    layout(
      title = paste("Violin Plot of", variable, "by Stroke"),
      xaxis = list(title = "Stroke"),
      yaxis = list(title = variable)
    )
  
  plot
}

# Generate individual plots
plot_age <- create_violin_plot(outlier_stroke_tb, variable = "age", stroke_variable = "stroke")
plot_glucose <- create_violin_plot(outlier_stroke_tb, variable = "avg_glucose_level", stroke_variable = "stroke")
plot_bmi <- create_violin_plot(outlier_stroke_tb, variable = "bmi", stroke_variable = "stroke")

# Combine plots into a grid layout (1 row, 3 columns)
combined_plot <- subplot(
  plot_age, plot_glucose, plot_bmi,
  nrows = 1, titleX = TRUE, titleY = TRUE
) %>%
  layout(title = "Violin Plots of Numerical Variables by Stroke")

# Display the combined plot
combined_plot
```

The histogram and violin charts have been grouped by stroke occurrence. This allows us to observe the distribution of the data for each variable in relation to the target variable `stroke`. Based on the charts, `age` and `avg_glucose_level` exhibit notable differences in the proportions of stroke cases across their value ranges. Stroke cases are more frequent at higher glucose levels and in the age range of 60 to 80 years.

The variable `age` shows a weak positive relationship with the target variable `stroke` (correlation coefficient of 0.24). `avg_glucose_level` shows an even weaker positive relationship with the target variable (correlation coefficient of 0.13). In contrast, `bmi` and `stroke` show no relationship with a correlation coefficient of 0.04.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Calculate correlation coefficients for each variable with stroke
cor_age <- cor(as.numeric(outlier_stroke_tb$age), as.numeric(outlier_stroke_tb$stroke), method = "pearson")
cor_glucose <- cor(as.numeric(outlier_stroke_tb$avg_glucose_level), as.numeric(outlier_stroke_tb$stroke), method = "pearson")
cor_bmi <- cor(as.numeric(outlier_stroke_tb$bmi), as.numeric(outlier_stroke_tb$stroke), method = "pearson")

# Create a correlation table
correlation_table <- data.frame(
  Variable = c("Age", "Average Glucose Level", "BMI"),
  Correlation_with_Stroke = round(c(cor_age, cor_glucose, cor_bmi), 2)
)

# Display the correlation table with kable
kable(correlation_table, caption = "Correlation Coefficients with Stroke")
```

### Categorical variables

The second part of our EDA examines the distribution of the categorical variables and their relationship with the target variable in a stacked bar chart:

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Define the variables and their labels for categories
var_categories <- list(
  "smoking_status" = c("Unknown", "Never smoked", "Formerly smoked", "Smokes"),
  "hypertension" = c("No", "Yes"),
  "heart_disease" = c("No", "Yes"),
  "ever_married" = c("No", "Yes"),
  "work_type" = c("Govt_job", "Never_worked", "Private", "Self-employed", "children")
)

# Filter data and select variables of interest
filtered_data <- outlier_stroke_tb %>%
  filter(gender != "Other") %>%
  select(stroke, names(var_categories))

# Convert stroke to a factor with labels
filtered_data <- filtered_data %>%
  mutate(stroke = factor(stroke, levels = c(0, 1), labels = c("No Stroke", "Stroke")))

# Pivot data to long format
long_data <- filtered_data %>%
  pivot_longer(cols = all_of(names(var_categories)), 
               names_to = "variable", values_to = "category")

# Compute the proportion of "Stroke" for each variable-category combination
order_data <- long_data %>%
  group_by(variable, category, stroke) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(variable, category) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  filter(stroke == "Stroke")

# For each variable, reorder categories by ascending proportion of Stroke
order_data <- order_data %>%
  group_by(variable) %>%
  arrange(prop, .by_group = TRUE) %>%
  mutate(cat_rank = row_number()) %>%
  ungroup()

# Compute a variable-level order by mean stroke proportion
var_order_data <- order_data %>%
  group_by(variable) %>%
  summarize(var_prop = mean(prop), .groups = "drop") %>%
  arrange(var_prop) %>%
  mutate(var_rank = row_number())

# Join category and variable ranks back to long_data
long_data <- long_data %>%
  left_join(order_data %>% select(variable, category, cat_rank), by = c("variable", "category")) %>%
  left_join(var_order_data %>% select(variable, var_rank), by = "variable")

# Create a combined label "Variable: Category"
long_data <- long_data %>%
  mutate(variable_category = paste(variable, category, sep = ": "))

# Arrange by var_rank (variable-level order), then cat_rank (category-level order)
long_data <- long_data %>%
  arrange(var_rank, cat_rank)

# Set factor levels according to this ordering
long_data$variable_category <- factor(long_data$variable_category, levels = unique(long_data$variable_category))

# Create the plot
p <- ggplot(long_data, aes(x = variable_category, fill = stroke)) +
  geom_bar(position = "fill") +
  labs(
    title = "Distribution Across Risk Factors",
    x = NULL,      # After flipping, becomes y-axis
    y = "Proportion",
    fill = "Stroke Status"
  ) +
  scale_fill_manual(values = c("No Stroke" = "grey50", "Stroke" = "#1f77b4")) +
  scale_y_continuous(labels = percent) + # Convert proportion to percentage
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  coord_flip()

# Convert to Plotly
combined_plot <- ggplotly(p) %>%
  layout(
    margin = list(t = 50),
    showlegend = TRUE,
    legend = list(title = list(text = "<b>Stroke Status</b>"))
  )

# Display the combined horizontal bar chart
combined_plot
```

Relevant differences in the proportions of stroke cases can be observed across the levels of the following categorical variables:

-   `smoking_status`: The proportion of stroke cases is highest among individuals who formerly smoked, compared to those who never smoked or currently smoke. This relationship could be influenced by age, as individuals who formerly smoked may be older on average. Current smokers show a slightly higher proportion of stroke cases than those who never smoked.
-   `hypertension`: Individuals with hypertension have a higher proportion of stroke cases compared to those without hypertension.
-   `ever_married`: Married individuals show a higher proportion of stroke cases compared to those who have never been married. This relationship might be influenced by age, as younger individuals are less likely to be married, and the risk of stroke typically increases with age.
-   `work_type`: The proportion of stroke cases varies across work types, with private sector employees and self-employed individuals exhibiting higher proportions of stroke cases than individuals in other work categories.
-   `heart disease`: Individuals with heart disease have a higher proportion of stroke cases compared to those without heart disease.

In contrast, the following variables appear to be not relevant, as there are no differences in the proportions of stroke cases across their levels (see Appendix):

-   `gender`

-   `Residence_type`

Based on the univariate analysis, `heart_disease`, `smoking_status`, `hypertension`, `ever_married`, and `work_type` could be potential candidates for further exploration and modeling among the categorical variables. These variables exhibit notable differences in the proportions of stroke cases across their levels, suggesting they may be important predictors of stroke risk. Among the numerical variables, `age` and `avg_glucose_level` show a weak to very weak positive relationship with the target variable `stroke`, suggesting they may also be important predictors of stroke risk. These predictors may even become more relevant when examined in a balanced dataset, which our initial hypotheses would suggest.

## Bivariate Analysis

In the bivariate analysis we examined correlations between our features. We used a correlation matrix to visualize the relationships between the numerical variables and the target variable `stroke`.

```{r, message = FALSE}
#| code-fold: true
#| code-summary: "Click to show code"

# Calculate the correlation matrix
cormat <- cor(numerical_data, use = "complete.obs")

# Get lower triangle of the correlation matrix
  get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
  }
  # Get upper triangle of the correlation matrix
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }

# Helper function to reorder the correlation matrix
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}

# Reorder the correlation matrix (assuming reorder_cormat and get_upper_tri are defined)
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)

# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)

# Create the base heatmap
gheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1,1), space = "Lab",
    name = "Pearson Correlation"
  ) +
  coord_fixed()

# Add correlation coefficients, adjust theme and legend, and add a centered top title
gheatmap + 
  geom_text(aes(Var2, Var1, label = round(value, 2)), color = "black", size = 4) +
  ggtitle("Correlation Matrix Heatmap") +
  theme(
    plot.title.position = "plot",     # Title positioned at the top of the plot area
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # Center the title
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "right"
  ) +
  guides(fill = guide_colorbar(
    barwidth = 1,
    barheight = 7,
    title.position = "top",
    title.hjust = 0.5
  ))

```

Our numerical features don't seem to be strongly correlated with each other. The strongest correlation is observed between `age` and `bmi` (0.324), which is not surprising, as BMI is often used as a measure of body fat and is known to increase with age. The correlation between the other numerical features is rather low. We can therefore safely assume independence of our numerical features. A further examination of the categorical features for collinearity will be done in the final data preparation in the next part.
