---
title: "_introduction"
format: html
server: shiny
---
# Introduction and Data Structure

## Project Goals

Describe the main goals of the project, including the motivation behind the research and the key questions you aim to answer.

## Research Questions

-   Question 1: ...
-   Question 2: ...
-   (Add as necessary)

### Adjustments or Refinements

Discuss any changes or refinements to the project scope or questions based on initial findings or data constraints.

## Data Source
The data utilized for this project originates from the Stroke Prediction Dataset available on Kaggle, which is a public platform that hosts a variety of datasets curated by data scientists and researchers. This specific dataset includes health-related variables that are relevant for predicting occurrences of strokes based on features like age, gender, and medical history. Unfortunately, we do not have more information on how the data was collected. The source is confidential and the use of it is only permitted for educational purposes.

## Data Structure

```{r}
#import of the dataset
stroke_db <- read.csv("../data/stroke_dataset.csv")

#saving as a tibble
stroke_tb <- as_tibble(stroke_db)
head(stroke_tb)
```

In total, there are 12 variables including our binary target variable `stroke`. 

Our dataset contains 5110 observations with 12 attributes. Here's a summary of the main features of the dataset:

1.  `id`:

-   Type: Integer (Continuous)
-   Description: A unique identifier for each individual in the dataset

2.  `gender`:

-   Type: Character
-   Values: "Male", "Female", or "Other"
-   Description: Gender of the individual

3.  `age`:

-   Type: Double
-   Description: Age of the individual

4.  `hypertension`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: Indicates whether the individual has hypertension (high blood pressure)

5.  `heart_disease`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: Indicates whether the individual has a heart disease

6.  `ever_married`:

-   Type: Character
-   Values: "No" or "Yes"
-   Description: Indicates whether the individual has ever been married

7.  `work_type`:

-   Type: Character
-   Values: "children", "Govt_job", "Never_worked", "Private", "Self-employed"
-   Description: Type of work the individual is engaged in

8.  `Residence_type`:

-   Type: Character
-   Values: "Rural" or "Urban"
-   Description: Indicates the living area of the individual

9.  `avg_glucose_level`:

-   Type: Double
-   Description: Average glucose level in the blood

10. `bmi`:

-   Type: Character (Originally), Double (After Preprocessing)
-   Description: Body Mass Index, a measure of body fat based on height and weight

11. `smoking_status`:

-   Type: Character
-   Values: "formerly smoked", "never smoked", "smokes", "Unknown"
-   Description: Indicates the smoking habits of the individual, where "Unknown" signifies that the smoking status information is not available

12. `stroke`:

-   Type: Integer (Binary)
-   Values: 0 (No), 1 (Yes)
-   Description: The target variable indicating whether the individual has had a stroke

```{r, warning=FALSE}
#|code-fold: true 
#|code-summary: "Click to show code"
stroke_tb <- read.csv("../data/stroke_dataset.csv") 
head(stroke_tb)
str(stroke_tb)
```
Relevant metadata:

-   Owner of the dataset: Federico Soriano Palacios (found thanks to the link to his LinkedIn on his     Kaggle profile)

-   Last update of the dataset was 4 years ago 
